---
title: "Influenza Deaths Forecast"
author: "Anurag Nepal"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Background

I wanted to forecast US pneumonia and influenza mortality as a time series problem. Weekly surveillance death counts have strong seasonality and ugly reporting artifacts, including the COVID-era spikes. The question was which modeling approach gives a usable 24-month forecast: ETS, ARIMA on raw versus cleaned data, and finally a hybrid that lets volatility vary over time.

# Setup

The analysis reads `National_Custom_Data.csv`, which is not in this repo. It is a custom data download from CDC FluView's pneumonia, influenza, and COVID-19 mortality surveillance (NCHS), National region. To reproduce: open the FluView mortality surveillance page, use the Downloads tool, and export the custom National-level dataset. The columns used here are `SEASON`, `WEEK`, `NUM.INFLUENZA.DEATHS`, `NUM.PNEUMONIA.DEATHS`, and `TOTAL.DEATHS`.

```{r}
# Load necessary libraries
library(forecast)
library(tseries)
library(dplyr)
library(lubridate)
library(knitr)
library(ggplot2)
library(astsa)
library(itsmr)
library(ltsa)
library(car)
```

# Method

## Loading and cleaning

The raw counts arrive comma-formatted, so the first job is parsing them into numbers. Flu seasons run week 40 to week 39, so I map each season/week pair to a calendar date, then aggregate to monthly level. The target is the percent of all deaths attributable to pneumonia and influenza. I keep two versions of the series: the raw one, and a cleaned one with outliers interpolated via `tsclean` for more robust modeling.

```{r}
# ========================================================
# DATA LOADING & CLEANING
# ========================================================
# Load Data
# National_Custom_Data.csv holds weekly influenza/pneumonia death counts from
# CDC FluView (NCHS mortality surveillance), National region custom download.
# It is not in this repo; place it next to this file to re-run the analysis.
raw_data <- read.csv("National_Custom_Data.csv", stringsAsFactors = FALSE)
head(raw_data)
# Clean Numerical Columns (Remove commas)
clean_col <- function(x) as.numeric(gsub(",", "", x))
raw_data$NUM.INFLUENZA.DEATHS <- clean_col(raw_data$NUM.INFLUENZA.DEATHS)
raw_data$NUM.PNEUMONIA.DEATHS <- clean_col(raw_data$NUM.PNEUMONIA.DEATHS)
raw_data$TOTAL.DEATHS <- clean_col(raw_data$TOTAL.DEATHS)

# Create Date Column
get_cal_year <- function(season, week) {
  start_year <- as.numeric(substr(season, 1, 4))
  if (week >= 40) return(start_year)
  else return(start_year + 1)
}
raw_data$Year <- mapply(get_cal_year, raw_data$SEASON, raw_data$WEEK)
raw_data$Date <- as.Date(paste(raw_data$Year, raw_data$WEEK, 1, sep="-"), format="%Y-%U-%u")

# Aggregate to Monthly Level
monthly_data <- raw_data %>%
  group_by(MonthDate = floor_date(Date, "month")) %>%
  summarise(
    Total_P_I = sum(NUM.INFLUENZA.DEATHS + NUM.PNEUMONIA.DEATHS, na.rm = TRUE),
    Total_All_Deaths = sum(TOTAL.DEATHS, na.rm = TRUE)
  ) %>%
  mutate(Percent_PI = (Total_P_I / Total_All_Deaths) * 100) %>%
  arrange(MonthDate) %>%
  filter(!is.na(MonthDate))

# Create Time Series Objects
start_yr <- year(min(monthly_data$MonthDate))
start_mo <- month(min(monthly_data$MonthDate))

# 1. Raw Data
ts_raw <- ts(monthly_data$Percent_PI, start = c(start_yr, start_mo), frequency = 12)

# 2. Cleaned Data (Outliers Interpolated for Robust Modeling)
ts_cleaned <- tsclean(ts_raw)
```

```{r}
summary(monthly_data$Percent_PI)
summary(ts_cleaned)
```

## Exploring the series

The raw plot shows the seasonal winter peaks plus the COVID-era spikes. After cleaning, the STL decomposition shows the seasonality clearly, and the ACF/PACF guide the ARIMA orders.

```{r}
# Plot Raw Data
plot(ts_raw, main="US P&I Mortality Rate (Raw Data)", ylab="% Deaths", col="darkblue", lwd=1.5)
grid()
plot(ts_cleaned, main="US P&I Mortality Rate (Cleaned Data)", ylab="% Deaths", col="darkblue", lwd=1.5)
grid()

# STL Decomposition (Using Cleaned Data to see clear seasonality)
decomp <- stl(ts_cleaned, s.window="periodic")
plot(decomp, main="STL Decomposition (Cleaned Data)")

# ACF and PACF
par(mfrow=c(1,2))
Acf(ts_cleaned, main="ACF (Cleaned Data)")
Pacf(ts_cleaned, main="PACF (Cleaned Data)")
par(mfrow=c(1,1))
```

## ETS baseline

First model: ETS with automatic error/trend/seasonality selection, fit on the cleaned series, with a 24-month forecast and residual diagnostics.

```{r}
# We use ETS (Error, Trend, Seasonality).
# model="ZZZ" allows the function to automatically select the best error/trend/season type

# Fit ETS Model
fit_ets <- forecast::ets(ts_cleaned, model="ZZZ") 

# Summary of chosen model
print(summary(fit_ets))

# Plot ETS Forecast
ets_fc <- forecast::forecast(fit_ets, h=24)

plot(ets_fc)
test(ets_fc$residuals)
```

## Checking stationarity

Before ARIMA, I ran ADF tests on the level, the seasonally differenced series, and the doubly differenced series to see how much differencing the data needs.

```{r}
# 1. Test on Level Data
print("--- ADF Test: Level ---")
print(aTSA::adf.test(ts_cleaned)) 

# 2. Test on Seasonal Difference (D=1)
ts_diff <- diff(ts_cleaned, lag=12)
print("--- ADF Test: Seasonal Difference (D=1) ---")
print(aTSA::adf.test(ts_diff))

# 3. Test on Double Difference (D=1, d=1)
# Often required if seasonal differencing alone doesn't remove the trend
ts_doubly_diff <- diff(ts_diff, lag=1)
print("--- ADF Test: Seasonal + First Difference (D=1, d=1) ---")
print(aTSA::adf.test(ts_doubly_diff))
```

## ARIMA bake-off

Three candidates, all fit with `auto.arima`: one on the raw data, one on the cleaned data, and one on the cleaned data with a log transform. The trace output is hidden to keep the document readable.

```{r arima_fits, results='hide'}
# Model 1: ARIMA on Raw Data
fit_raw <- auto.arima(ts_raw, seasonal=TRUE, stepwise=FALSE, approximation=FALSE, trace=TRUE)

# Model 2: ARIMA on Cleaned Data (No Transformation)
fit_clean <- auto.arima(ts_cleaned, seasonal=TRUE, stepwise=FALSE, approximation=FALSE, trace=TRUE)

# Model 3: ARIMA on Cleaned Data + Log Transformation
fit_log_clean <- auto.arima(ts_cleaned, seasonal=TRUE, lambda=0, approximation=FALSE, trace=TRUE)
```

```{r}
# Function to extract metrics
get_accuracy <- function(model, name) {
  acc <- accuracy(model)
  return(data.frame(
    Model = name,
    AICc = round(model$aicc, 2),
    RMSE = round(acc[2], 4),
    MAPE = round(acc[5], 4)
  ))
}

# Combine into a table
comparison_table <- rbind(
  get_accuracy(fit_raw, "ARIMA (Raw Data)"),
  get_accuracy(fit_clean, "ARIMA (Cleaned Data)"),
  get_accuracy(fit_log_clean, "ARIMA (Cleaned + Log)")
)

kable(comparison_table, caption = "Model Accuracy Comparison")
```

Residual diagnostics for all three fits:

```{r}
test(fit_raw$residuals)
test(fit_clean$residuals)

test(fit_log_clean$residuals)

```

The log-transformed model on the cleaned series became the final forecasting model. Below are all three forecasts for comparison, then the chosen 24-month forecast with its values tabulated.

```{r}
# Generate Forecasts
fc_raw <- forecast::forecast(fit_raw, h=24)
fc_clean <- forecast::forecast(fit_clean, h=24)
fc_log_clean  <- forecast::forecast(fit_log_clean, h=24, biasadj=TRUE)

plot(fc_raw, main="Forecast: Raw Model (No Log Transform)", ylab="% Deaths", fcol="blue")
lines(ts_raw, col="black")
# PLOT 1: Cleaned Data Model
plot(fc_clean, main="Forecast: Cleaned Model (No Log Transform)", ylab="% Deaths", fcol="blue")
lines(ts_raw, col="black")


plot(fc_log_clean, 
     main="24-Month Forecast: US P&I Mortality (Log-Transformed+ Cleaned)", 
     ylab="% Deaths", 
     xlab="Year",
     fcol="#b00000",       # Dark Red
     flty=1, lwd=2,
     shadecols=c("#d9d9d9", "#bdbdbd")
     )        # Focus on realistic range

# Overlay Historical Data
lines(ts_raw, col="black", lwd=1.5)

legend("topleft", 
       legend=c("Historical (Raw)", "Forecast (Log-ARIMA)", "95% CI"),
       col=c("black", "#b00000", "#bdbdbd"), 
       lty=c(1, 1, 1), lwd=c(1.5, 2, 8), bty="n")
```

```{r}
# Extract forecast values into a readable dataframe
fc_values <- data.frame(
  Date = seq(from = max(monthly_data$MonthDate) %m+% months(1), by = "month", length.out = 24),
  Forecast = round(fc_log_clean$mean, 2),
  Lower_95 = round(fc_log_clean$lower[,"95%"], 2),
  Upper_95 = round(fc_log_clean$upper[,"95%"], 2)
)

kable(fc_values, caption = "24-Month Forecast Values (Log-ARIMA)")
```

## Hybrid ARIMA plus GARCH

The ARIMA mean forecast assumes constant variance, but mortality spikes cluster. So I fit a GARCH(1,1) with Student-t errors on the raw ARIMA residuals, deliberately using the raw fit's residuals to keep the real spikes in the volatility estimate. The conditional volatility plot shows when uncertainty was highest.

```{r}
library(rugarch)
# 1. Extract Residuals from the best ARIMA fit
best_residuals <- residuals(fit_raw) # Using Raw ARIMA residuals to capture the real spikes

# 2. Define GARCH(1,1) Specification
garch_spec <- ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model = list(armaOrder = c(0, 0), include.mean = FALSE),
  distribution.model = "std" 
)

# 3. Fit GARCH
garch_fit <- ugarchfit(spec = garch_spec, data = best_residuals)

# 4. Diagnostics
print(garch_fit) 

# Plot Conditional Volatility (Sigma)
volatility <- sigma(garch_fit)
plot(volatility, type="l", col="darkred", 
     main="Conditional Volatility (Sigma) - GARCH(1,1)",
     ylab="Estimated Volatility", xlab="Time")
grid()
```

```{r}
# Extract Standardized Residuals
std_resid <- residuals(garch_fit, standardize=TRUE)

# Diagnostic Tests (Ljung-Box, etc.)
test(as.numeric(std_resid))
```

# Results

The final hybrid forecast combines the ARIMA mean path with GARCH-driven 95% intervals: the interval width at each horizon comes from the forecasted conditional volatility rather than a constant variance assumption.

```{r}
# 8. Final Hybrid Forecasting

forecast_horizon <- 24

# 1. ARIMA Forecast (Mean)
fc_arima <- forecast::forecast(fit_raw, h=forecast_horizon)

# 2. GARCH Forecast (Variance/Sigma)
garch_fc <- ugarchforecast(garch_fit, n.ahead = forecast_horizon)
sigma_forecast <- sigma(garch_fc)

# 3. Create Hybrid Dataframe
start_date <- max(monthly_data$MonthDate)
future_dates <- seq(from = start_date, by = "month", length.out = forecast_horizon + 1)[-1]

hybrid_df <- data.frame(
  Date = future_dates,
  Mean_Forecast = as.numeric(fc_arima$mean),
  Sigma_GARCH = as.numeric(sigma_forecast)
)

# Calculate 95% Confidence Intervals
hybrid_df$Lower_95 <- hybrid_df$Mean_Forecast - (1.96 * hybrid_df$Sigma_GARCH)
hybrid_df$Upper_95 <- hybrid_df$Mean_Forecast + (1.96 * hybrid_df$Sigma_GARCH)

# 4. Final Plot

library(lubridate)
forecast_time_axis <- decimal_date(hybrid_df$Date)

# 2. Define Plot Limits
start_plot <- 2018
end_plot   <- max(time(ts_raw)) + (24/12) 

# 3. Plot
plot(ts_raw, 
     main="24-Month Forecast: Hybrid ARIMA + GARCH(1,1)", 
     ylab="% Deaths", xlab="Year",
     xlim=c(start_plot, end_plot), 
     ylim=c(min(hybrid_df$Lower_95), max(ts_raw)),
     col="black", lwd=1)

# 4. Add Lines using the converted time axis
lines(forecast_time_axis, hybrid_df$Mean_Forecast, col="blue", lwd=2)
lines(forecast_time_axis, hybrid_df$Upper_95, col="red", lty=2)
lines(forecast_time_axis, hybrid_df$Lower_95, col="red", lty=2)

legend("topleft", 
       legend=c("Historical", "Forecast Mean", "GARCH 95% CI"),
       col=c("black", "blue", "red"), 
       lty=c(1, 1, 2), lwd=c(1, 2, 1), bty="n")

```

# Takeaway

The cleaning step was the difference between a model that behaves and one that does not: comma-formatted counts and reporting artifacts had to be handled before any ARIMA would fit sensibly. On the cleaned series, the log-transformed ARIMA gave the final 24-month mean forecast, and layering GARCH(1,1) volatility on the residuals produced honest 95% intervals that widen where history was volatile. The hybrid keeps the seasonal structure from ARIMA while admitting that uncertainty itself clusters in time.
